Model Finding for Recursive Functions in SMT\thanks{ This research is partially supported by the Inria technological development action ``Contre-exemples Utilisables par Isabelle et Coq'' (CUIC). }

### 2015-05-05 tex/main.tex
tuned whitespaces
more

\documentclass[runningheads,a4paper]{llncs}
\usepackage[T1]{fontenc}
\usepackage[scaled=.82]{beramono}
\usepackage[scaled=.86]{helvet}
\usepackage{mathptmx}
\usepackage{listings}
\usepackage{colonequals}
%\usepackage{mathpartir}
\usepackage{xspace}
\usepackage{amssymb}
\setcounter{tocdepth}{3}
\usepackage{graphicx}
\usepackage{stmaryrd}
\usepackage{cite}
\usepackage{hyperref}
\usepackage{framed}

@@

@@

\newcommand\xend{{\hfill$\scriptstyle\blacksquare$}}

@@

@@

In this paper, we consider \emph{annotated quantifed formulas} of the form
$\forallf{a} \vec x.\; \varphi$, where the annotation~$a$ is either
an uninterpreted function $\con{f} \in \sfuns{\Sigma}$ or the special symbol
$\fnull$.
Given $\con{f} : \tau_1 \times \ldots \times \tau_n \rightarrow \tau$,
an annotated
quantified formula $\forallf{\con{f}} \vec x.\; \varphi$ is a \emph{function definition}
(\,\emph{for $\con{f}$}\vthinspace) if $\vec x$ is a tuple of variables of type
$\tau_1, \ldots, \tau_n$ and $\varphi$ is a (quantifier-free) formula equivalent to
$\con{f}( \vec x ) \teq t$ for some $t$. We write $\forall \vec
x.\; \varphi$ as shorthand for $\forallf{\fnull} \vec x.\; \varphi$, and $\exists
\vec x. \varphi$ as shorthand for $\neg \forall \vec x.\; \neg \varphi$.
For a formula $\varphi$, 
we write $\varphi[\vec t/\vec x]$ to denote the result of replacing all occurrences of $\vec x$ with $\vec t$.

@@

\begin{definition}
Given a set of function definitions $\DDD = \{ \forallf{f_1} \vec x.\; \varphi_1, \ldots, \forallf{f_n} \vec x.\; \varphi_n \}$,
a formula $\varphi$ is \emph{closed under function expansion with respect to $\DDD$} if
\begin{eqnarray}
\varphi \models_T \displaystyle\bigwedge\limits_{i=1}^n \{ \varphi_i[ \vec t / \vec x ] \mid f_i( \vec t ) \in \terms( \varphi ) \}
\end{eqnarray}
The set $\DDD$ is \emph{admissible} if whenever there exists a model for $\varphi$ that is closed under function expansion,
there exists a model for $\varphi \wedge \forallf{f_1} \vec x.\; \varphi_1 \wedge \ldots \wedge \forallf{f_n} \vec x.\; \varphi_n$.
\end{definition}

@@

In more detail, 
suppose our signature $\Sigma$ contains a set of defined function symbols $\sfundefs{\Sigma} \subseteq \sfuns{\Sigma}$.
We extend the signature of $\Sigma$ to a signature $\extendsig{\Sigma}$ containing the following additional definitions.
For each $\con{f} : \tau_1 \times \ldots \times \tau_n \rightarrow \tau \in \sfundefs{\Sigma}$,
the signature $\extendsig{\Sigma}$ includes
\begin{itemize}
\item an uninterpreted \emph{abstract type} $\fargsort{\con{f}}$;
\item $n$ uninterpreted \emph{concretization functions} $\fargx{\con{f}}{1} : \fargsort{\con{f}} \rightarrow \tau_1$, \ldots, $\fargx{\con{f}}{n} : \fargsort{\con{f}} \rightarrow \tau_n$.
\end{itemize}
%The interpretation of uninterpreted sort $\fargsort{\con{f}}$ will denote the elements (tuples) on which the function $\con{f}$ is applied.
%The role of the uninterpreted functions $\farg{\con{f}}^1$, $\ldots$, $\farg{\con{f}}^n$ will be discussed more in the following.

@@

\begin{theorem}
If $\varphi$ is a\/ $\Sigma$-formula that is in definitional form with respect to\/
$\sfundefs{\Sigma}$
and the set of function definitions\/ $\DDD$ corresponding to functions in
$\sfundefs{\Sigma}$ is admissible, then\/ $\varphi$ and\/ $\conv( \varphi )$ are
equisatisfiable.
\end{theorem}
\begin{proof}
(Sketch)  First, we show that if $\varphi$ is satisfied by $\Sigma$-interpretation $\I$, then $\conv( \varphi )$ is satisfied by a $\extendsig{\Sigma}$-interpretation $\J$.
Let $\J$ be the $\extendsig{\Sigma}$-interpretation that interprets all types $\tau \in \ssorts{\Sigma}$ as $\tau^\I$,
all functions $\con{f} \in \sfuns{\Sigma}$ as $\con{f}^\I$,
and for each function $\con{f}$ of type $\tau_1 \times \ldots \tau_n \rightarrow \tau$ in $\sfundefs{\Sigma}$, 
interprets $\fargsort{\con{f}}$ as a tuple of values of type $\tau_1^\I \times \ldots \times \tau_n^\I$,
and each $\fargx{\con{f}}{i}$ as the $i^{th}$ projection on such tuples for each $i = 1, \ldots, n$.
Note that $\J$ satisfies \emph{every} constraint of the form $\exists a : {\fargsort{\con{f}}}. ( \vecfarg{\con{f}}( a ) \teq \vec t )$,
since by our construction of $\J$ there exists an $a$ such that $a^\J = \vec t^\J$.
By case analysis it follows that $\conv_0$ maintains the invariant that
if $\conv_0( t\typ{\Bool}, \ppos )$ returns $( r\typ{\Bool}, D )$,
then $\I$ satisfies $t$ if and only if $\J$ satisfies $r$.
Thus, we conclude $\J$ is a model of $\conv( \varphi )$.

@@

Since there exists a model for $\psi$ that is closed under function expansion with respect to $\DDD$, 
then since $\DDD$ is admissible,
by definition there exists a model for $\psi \wedge \DDD$, which entails $\DDD \wedge \varphi_0$, which is $\varphi$.
\end{proof}

Let us revisit the translated formula~(\ref{eq:ex-after}) from Example~\ref{ex:translation}.
This formula is unsatisfiable if and only if formula~(\ref{eq:ex-before}) is unsatisfiable since $\fargsort{f}$ can be interpreted as $\sortint$,
and $\farg{f}$ can be interpreted as the identity function.
Conversely, we claim formula~(\ref{eq:ex-after}) is satisfiable if and only if formula~(\ref{eq:ex-before}) is satisfiable,
noting that the set $\{ \forallf{\con{s}} x\typ{\sortint}.\;\, \lite\bigl( x \leq 0, \con{s}(x) \teq 0, \con{s}( x ) \teq x + \con{s}( x-1 ) \bigr) \}$ is admissble.
As stated in the proof of the previous theorem, 
the reason is that any model of formula~(\ref{eq:ex-after}) must satisfy a formula that is closed under function expansion,
hence implying that formula~(\ref{eq:ex-before}) has a model.
In more detail, let $\varphi_s$ be the formula $\forallf{\con{s}} x.\;\, \lite\bigl( x \leq 0, \con{s}(x) \teq 0, \con{s}( x ) \teq x + \con{s}( x-1 ) \bigr)$.
The rightmost conjunct 
$\exists a : {\fargsort{\con{s}}}. \farg{\con{s}}( a ) \teq \con{c}$ ensures 
that any model of formula~(\ref{eq:ex-after}) satisfying $\con{s}( \con{c} ) > 100$ also satisfies $\varphi_{\con{s}}[\con{c}/x]$.
Similarly, the nested existential $\exists a.\;\, \farg{\con{s}}( a ) \teq \farg{\con{s}}( y )-1$
ensures that this constraint is enforced recursively.
Given a model $\I$ of formula~(\ref{eq:ex-after}) that interprets $c$ as $14$, 
it is easy to see that $\I$ must satisfy the formula:
\begin{eqnarray} \label{eq:fun-closure}
\begin{split}
\psi := \con{s}( c ) > 100 \wedge & \displaystyle\bigwedge\limits_{i=0}^{13} ( \neg (c-i \leq 0) \wedge \con{s}( c-i ) \teq i + \con{s}( c-i-1 ) ) \\
                            & \wedge c-14 \leq 0 \wedge \con{s}( c-14 ) \teq 0
\end{split}
\end{eqnarray}
Notice that $\psi$ is closed under function expansion, 
since it contains the $s$-applications that are equivalent to $\con{s}( c-i )$ for $i = 0, \ldots, 14$, and
entails $\varphi_s[c-i/x]$ for $i = 0, \ldots, 14$.
Since $\varphi_s$ is an admissible definition,
this implies there exists a model for $\psi \wedge \varphi_s$,
which implies there exists a model for formula~(\ref{eq:ex-before}).

>>>>

\documentclass[runningheads,a4paper]{llncs}
\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[scaled=.82]{beramono}
\usepackage[scaled=.86]{helvet}
\usepackage{mathptmx}
\usepackage{listings}
\usepackage{colonequals}
%\usepackage{mathpartir}
\usepackage{xspace}
\setcounter{tocdepth}{3}
\usepackage{graphicx}
\usepackage{stmaryrd}
\usepackage{cite}
\usepackage[
   a4paper,
   pdftex,
   pdftitle={Model Finding for Recursive Functions in SMT},
   pdfauthor={Andrew Reynolds, Jasmin Christian Blanchette, and Cesare Tinelli},
   pdfkeywords={},
   pdfborder={0 0 0},
   draft=false,
   bookmarksnumbered,
   bookmarks,
   bookmarksdepth=2,
   bookmarksopenlevel=2,
   bookmarksopen]{hyperref}
\usepackage{framed}

@@

\spnewtheorem*{proofsketch}{Proof sketch}{\itshape}{\rmfamily}

@@

@@

\newcommand\xend{{\hfill$\scriptstyle\blacksquare$}}
\renewcommand\qed{{\hfill$\squareforqed$}}

@@

In this paper, we consider \emph{annotated quantifed formulas} of the form
$\forallf{a} \vec x.\; \varphi$, where the annotation~$a$ is either
an uninterpreted function $\con{f} \in \sfuns{\Sigma}$ or the special symbol
$\fnull$.
Given $\con{f} : \tau_1 \times \cdots \times \tau_n \rightarrow \tau$,
an annotated
quantified formula $\forallf{\con{f}} \vec x.\; \varphi$ is a \emph{function definition}
(\,\emph{for $\con{f}$}\vthinspace) if $\vec x$ is a tuple of variables of type
$\tau_1, \ldots, \tau_n$ and $\varphi$ is a (quantifier-free) formula equivalent to
$\con{f}( \vec x ) \teq t$ for some $t$. We write $\forall \vec
x.\; \varphi$ as shorthand for $\forallf{\fnull} \vec x.\; \varphi$, and $\exists
\vec x. \varphi$ as shorthand for $\neg \forall \vec x.\; \neg \varphi$.
For a formula $\varphi$,
we write $\varphi[\vec t/\vec x]$ to denote the result of replacing all occurrences of $\vec x$ with $\vec t$.

@@

\begin{definition}
Given a set of function definitions $\DDD = \{ \forallf{f_1} \vec x.\; \varphi_1, \ldots, \forallf{f_n} \vec x.\; \varphi_n \}$,
a formula $\varphi$ is \emph{closed under function expansion with respect to $\DDD$} if
\begin{eqnarray}
\varphi \models_T \bigwedge_{i=1}^n \{ \varphi_i[ \vec t / \vec x ] \mid f_i( \vec t ) \in \terms( \varphi ) \}
\end{eqnarray}
The set $\DDD$ is \emph{admissible} if whenever there exists a model for $\varphi$ that is closed under function expansion,
there exists a model for $\varphi \wedge \forallf{f_1} \vec x.\; \varphi_1 \wedge \ldots \wedge \forallf{f_n} \vec x.\; \varphi_n$.
\end{definition}

@@

In more detail,
suppose our signature $\Sigma$ contains a set of defined function symbols $\sfundefs{\Sigma} \subseteq \sfuns{\Sigma}$.
We extend the signature of $\Sigma$ to a signature $\extendsig{\Sigma}$ containing the following additional definitions.
For each $\con{f} : \tau_1 \times \cdots \times \tau_n \rightarrow \tau \in \sfundefs{\Sigma}$,
the signature $\extendsig{\Sigma}$ includes
\begin{itemize}
\item an uninterpreted \emph{abstract type} $\fargsort{\con{f}}$;
\item $n$ uninterpreted \emph{concretization functions} $\fargx{\con{f}}{1} : \fargsort{\con{f}} \rightarrow \tau_1$, \ldots, $\fargx{\con{f}}{n} : \fargsort{\con{f}} \rightarrow \tau_n$.
\end{itemize}
%The interpretation of uninterpreted sort $\fargsort{\con{f}}$ will denote the elements (tuples) on which the function $\con{f}$ is applied.
%The role of the uninterpreted functions $\farg{\con{f}}^1$, $\ldots$, $\farg{\con{f}}^n$ will be discussed more in the following.

@@

\begin{theorem}\label{thm:equi}
If $\varphi$ is a\/ $\Sigma$-formula that is in definitional form with respect to\/
$\sfundefs{\Sigma}$
and the set of function definitions\/ $\DDD$ corresponding to functions in
$\sfundefs{\Sigma}$ is admissible, then\/ $\varphi$ and\/ $\conv( \varphi )$ are
equisatisfiable.
\end{theorem}
\begin{proofsketch}
First, we show that if $\varphi$ is satisfied by $\Sigma$-interpretation $\I$, then $\conv( \varphi )$ is satisfied by a $\extendsig{\Sigma}$-interpretation $\J$.
Let $\J$ be the $\extendsig{\Sigma}$-interpretation that interprets all types $\tau \in \ssorts{\Sigma}$ as $\tau^\I$,
all functions $\con{f} \in \sfuns{\Sigma}$ as $\con{f}^\I$,
and for each function $\con{f}$ of type $\tau_1 \times \cdots \tau_n \rightarrow \tau$ in $\sfundefs{\Sigma}$,
interprets $\fargsort{\con{f}}$ as a tuple of values of type $\tau_1^\I \times \cdots \times \tau_n^\I$,
and each $\fargx{\con{f}}{i}$ as the $i^{th}$ projection on such tuples for each $i = 1, \ldots, n$.
Note that $\J$ satisfies \emph{every} constraint of the form $\exists a : {\fargsort{\con{f}}}. ( \vecfarg{\con{f}}( a ) \teq \vec t )$,
since by our construction of $\J$ there exists an $a$ such that $a^\J = \vec t^\J$.
By case analysis it follows that $\conv_0$ maintains the invariant that
if $\conv_0( t\typ{\Bool}, \ppos )$ returns $( r\typ{\Bool}, D )$,
then $\I$ satisfies $t$ if and only if $\J$ satisfies $r$.
Thus, we conclude $\J$ is a model of $\conv( \varphi )$.

@@

Since there exists a model for $\psi$ that is closed under function expansion
with respect to $\DDD$, then since $\DDD$ is admissible, by definition there
exists a model for $\psi \wedge \DDD$, which entails $\DDD \wedge \varphi_0$,
which is $\varphi$.
\qed
\end{proofsketch}

\begin{example}
Let us revisit the translated formula~(\ref{eq:ex-after}) from
Example~\ref{ex:translation}. This formula is unsatisfiable if and only if
formula~(\ref{eq:ex-before}) is unsatisfiable since $\fargsort{f}$ can be
interpreted as $\sortint$, and $\farg{f}$ can be interpreted as the identity
function. Conversely, we claim formula~(\ref{eq:ex-after}) is satisfiable if
and only if formula~(\ref{eq:ex-before}) is satisfiable, noting that the set
$\{ \forallf{\con{s}} x\typ{\sortint}.\;\, \lite\bigl( x \leq 0, \con{s}(x)
\teq 0, \con{s}( x ) \teq x + \con{s}( x-1 ) \bigr) \}$ is admissble.
As stated
in the proof of Theorem~\ref{thm:equi}, the reason is that any model of
formula~(\ref{eq:ex-after}) must satisfy a formula that is closed under
function expansion, hence implying that formula~(\ref{eq:ex-before}) has a
model. In more detail, let $\varphi_\con{s}$ be the formula $\forallf{\con{s}} x.\;\,
\lite\bigl( x \leq 0, \con{s}(x) \teq 0, \con{s}( x ) \teq x + \con{s}( x-1 )
\bigr)$. The rightmost conjunct $\exists a : {\fargsort{\con{s}}}.
\farg{\con{s}}( a ) \teq \con{c}$ ensures that any model of
formula~(\ref{eq:ex-after}) satisfying $\con{s}( \con{c} ) > 100$ also
satisfies $\varphi_{\con{s}}[\con{c}/x]$. Similarly, the nested existential
$\exists a.\;\, \farg{\con{s}}( a ) \teq \farg{\con{s}}( y )-1$ ensures that
this constraint is enforced recursively. Given a model $\I$ of
formula~(\ref{eq:ex-after}) that interprets $\con{c}$ as $14$, it is easy to see that
$\I$ must satisfy the following formula $\psi$:
%
\begin{equation} \label{eq:fun-closure}
\begin{split}
& \con{s}( \con{c} ) > 100 \wedge %\displaystyle
  \bigwedge\nolimits_{\vthinspace i=0}^{\!13} \bigl( \neg (\con{c}-i \leq 0) \wedge \con{s}( \con{c}-i ) \teq i + \con{s}( \con{c}-i-1 ) \bigr) \\[-\jot]
{\wedge}\; & \con{c}-14 \leq 0 \wedge \con{s}( \con{c}-14 ) \teq 0
\end{split}
\end{equation}
%
This formula is closed under function expansion,
since it contains the $\con{s}$ applications corresponding to
$\con{s}( \con{c}-i )$ for $i = 0, \ldots, 14$. The formula also
entails $\varphi_\con{s}[\con{c}-i/x]$ for $i = 0, \ldots, 14$.
Since $\varphi_\con{s}$ is an admissible definition,
there must exist a model for $\psi \wedge \varphi_\con{s}$,
which means that there exists a model for formula~(\ref{eq:ex-before}).\xend
\end{example}
